---
title: "FTLE_Example"
author: "Taiki Sakai"
date: "2023-06-09"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Calculatin and Plotting FTLE

This document is a guide to calculating and plotting FTLEs based on the
online Restoration and TRACE tools.

First source required helper functions. 

```{r, message=FALSE}
library(here)
# sourcing Taiki's functions
source(here('FTLEfunctions.R'))
# and James' HFRADAR download function
source(here('hfRadar_Download.R'))
```
### Load GPS Data for Study Area

Starting point for analysis is to load some GPS data for the area we
want to study. Default expectation is that these are from our drifting
buoys, but the only real requirement is a dataframe with columns "UTC",
"Latitude", and "Longitude". Can read in multiple GPS files and combine
them in to a single dataframe.

```{r}
library(readr)
# this can be a vector of multiple
gpsFiles <- c(here('ADRIFT_067_GPS.csv'))
gps <- bind_rows(lapply(gpsFiles, read_csv))
```

### Download Current Data

Next step is to download the HFRADAR or HYCOM data to use for
calculating the FTLE. For HYCOM data we will download using the
`matchEnvData` function from PAMmisc, for HFRADAR data we will use
`gpsToHfradarDownload`, a wrapper around James Fahlbusch's
`hfRadar_Download`.

First setup a couple parameters for our data download. `intDays` is the
number of integration days we want to use for the FTLE calculation.
`coordBuffer` is how far around our GPS data we want to download
environmental data (decimal degrees). For example, if our GPS data runs
from Latitude 32 to 34, `coordBuffer = 1.5` will download data from
Latitude 30.5 to 35.5.

```{r}
intDays <- 2
coordBuffer <- 1.5
```

#### HYCOM

If using HYCOM data, we need to add a "Depth" column to our data with
value 0 so that we don't needlessly download all the depth levels
available in HYCOM.

```{r, eval=FALSE}
# Hycom
library(PAMmisc)
gps$Depth <- 0
currentNc <- 'ADRIFT067_HYCOM.nc'
gps <- matchEnvData(gps, nc=PAMmisc::hycomList, fileName = currentNc, var=c('water_u', 'water_v'), timeout=360,
                    buffer = c(coordBuffer, coordBuffer, intDays * 24 * 3600))
```

#### HFRADAR

If using HFRADAR data we don't need the "Depth" information because
HFRADAR only has surface currents.

```{r, eval=FALSE}
# HFRADAR
currentNc <- gpsToHfradarDownload(gps, days=intDays, buffer=coordBuffer, name='ADRIFT067')
```

Now we need to upload this `currentNc` file to the THREDDS server so
that we can run it through the restoration tool
(<http://transport.me.berkeley.edu/restore/>).

I've created a helper function to handle this FTP upload to the Berkeley
transport FTP server, but it requires password login information. For
safekeeping, that info is stored in a file secrets.yaml that we will
read in. We'll upload these to ADRIFT/Raw.

### Upload to THREDDS and Restoration

```{r, eval=FALSE}
library(yaml)
pw <- read_yaml(here('secrets.yaml'))
transportFtpUpload(currentNc, pw=pw$trace_pw, folder='ADRIFT/Raw')
```

The function should return "OK", and then your file should be visible
here:
<http://transport.me.berkeley.edu/thredds/catalog/public/Goldbogen-Lab/ADRIFT/Raw/catalog.html>
Click on your uploaded file, then click on the link for "OPENDAP". Copy
the full link in the "Data URL" field, and paste it into the "Enter URL"
field of the Restore tool. Hit "Scan Data" and ensure that it gives a
positive response.

Settings to check:

**Time:**

-   Time Interval - Check Process whole dataset times

**Domain:**

-   Specify Bounds - Check Use whole dataset bounds

**Generate Ensembles:**

-   Nothing checked

**Restore Data:**

-   Detect Domain - Check Concave hull around available points, enter
    10km.
-   Detect Land - Check Detect and Exclude Land
-   Refine Grid - Refinement level 1

Hit **Restore**!

Once it has finished running, hit **Download**, then move that NetCDF
file back to this project directory. I recommend re-naming it to have
the same name as the original .nc file, but with '\_restored' added at
the end. Then we'll use the FTP function to upload the restored file to
the "ADRIFT/Restored" folder.

### Upload to THREDDS and TRACE

```{r, eval=1}
# change to whatever you renamed your restored nc file to
restoredNc <- 'ADRIFT067_HYCOM_restored.nc'
transportFtpUpload(restoredNc, pw = pw$trace_pw, folder='ADRIFT/Restored')
```

That function should again return "OK", and your file should be visible
here:
<http://transport.me.berkeley.edu/thredds/catalog/public/Goldbogen-Lab/ADRIFT/Restored/catalog.html>
Same as before OPENDAP -\> copy "Data URL", paste that in the "Enter
URL" field of the TRACE tool
(<http://transport.me.berkeley.edu/trace/>).
Hit "Scan Data" and ensure it gives a positive response. 

Settings for this tool are slightly more complicated, I created a helper function
to report the proper number of values for some of the settings, 
`traceSettings`. Its inputs are the restored NetCDF file, `days`, the integration
days we assigned earlier, and `increment`, which is how far apart (hours) we want
each FTLE frame to be. Lower values of `increment` will take more total calculation
time, 4 might be a reasonable starting place.

```{r}
# note the reported Lat, Lon, and Times values for later
traceSettings(restoredNc, days=intDays, increment=4)
```

**Coordinate system**

* Check Cartesian

**Additional data**

* Check Include land mask variable in input data, and check Apply
free-slip boundary condition near land.

**Time**

* Initial Time - auto-fills from "Scan Data" - see note in "Time Direction" below
* Initialization type - Mutliple initial times
* Number of initial times - the "Times" reported from `traceSettings`
* Initial times increment - same as `increment` used earlier
* Integration duration - same as `intDays` parameter you entered above
* Time direction - forward for **repelling** FTLE, check backward for **attracting**
FTLE. **NOTE** if you check backward, add time to **Initial time** equal to
**Integration duration**

**Tracers**

* Select Uniform grid points
* Number of tracers along longitude - "Lon" value from `traceSettings`
* Number of tracers along latitude - "Lat" value from `traceSettings`
* Check Use dataset bounds

**Solver**

* Leave all as default

**Output**

* Output Variables - Check output trajectory coordinates at observation times
* Output Variables - Check Compute FTLE on trajectory
* Array structure - Select Grid-indexed arrays
* Visualization - Select none, I ran into errors trying to visualize in the web app

Hit **Run**! This will take a while, but I've noticed that the web app can lose connection
(and thus not finish calculations) if you leave your computer for too long. I've made a habit
of checking on the calculations every couple minutes to make sure its still running. 

Once finished hit download and move the file back to this project directory. I recommend 
renaming it to have "forward" or "backward" in the name depending on the time direction selected.

### Working With FTLE Output

Now we can read the FTLE data in to R with helper function `traceToDf`. This needs to know if it was
run forward or backward, and the number of integration days (these are used to properly assign times - 
times for each FTLE frame are set to be the middle of the integration period).

```{r, cache=TRUE}
ftleFile <- here('AD67_hycom_forward.nc')
# this can take a while and creates a very large dataframe
ftle <- traceToDf(ftleFile, backward=FALSE, days=intDays)
```

This dataframe contains data for *all* time frames, to make a single plot we want to look 
at one at a time, so we can filter by the `time` column. Here lets plot the first time frame.

```{r}
plotFTLE(ftle[ftle$time == unique(ftle$time)[1], ], gps=gps)
```

By default this uses the full extent of lat/long ranges in the GPS and FTLE data, we can
restrict the plot range with `xlim` and `ylim` parameters.

```{r}
plotFTLE(ftle[ftle$time == unique(ftle$time)[8], ], gps=gps,
         xlim = c(-124, -123), ylim=c(37, 38.5))
```

You can also create a GIF of the full time duration with the `plotFTLEGIF` function, 
this takes a bit of time to run.

```{r, eval=FALSE}
plotFTLEGIF(ftle, gps=gps, title='ADRIFT_067', progress=TRUE, file='ADRIFT_067_forward.gif')
```

You can also create side-by-side plots of the forward & backward FTLEs if you load in
both files and combine them into a single dataframe.

```{r, eval=FALSE}
ftleBack <- traceToDf('AD67_hycom_backward.nc', backward=TRUE, days=intDays)
ftleAll <- bind_rows(ftle, ftleBack)
plotFTLEGIF(ftleAll, gps=gps, title='ADRIFT_067', progress=TRUE, file='ADRIFT_067_both.gif')
```
